---
title: "CITE-seq based plots for Gearty et al. (2021)"
author: "Friederike Dündar | Weill Cornell Medicine"
date: "2021"
output:
    pdf_document:
        toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, cache=FALSE}
library(SingleCellExperiment)
library(magrittr); library(data.table)
library(ggplot2); theme_set(theme_bw(base_size = 14))
library(patchwork)
source("src/utils_SCE.R")
source("src/load_data_from_box.R")
linrary("src/marker_gene_extract_and_plot.R")
library(clusterProfiler)
```

These plots are based on scRNA-seq/CITE-seq data obtained from sorted CD8+ T cells from pancreatic lymph nodes (pLN).
The data was obtained from three technical replicates (`Sample`): "pLN_1", "pLN_3", "pLN_4" that correspond to three different library prep and sequencing runs.
Each of the experiments had CD8 T cells from multiple mice, whose identity was derived via antibody-based sample tagging and can be found in the `SingleCellExperiment`'s `colData` (meta data) via the `Mouse` entry.
CD44, CD62L etc. correspond to the normalized CITE-seq values for the respective antibodies

The SingleCellExperiment object and all other Robjects needed to recreate this document are stored in the Box; the [BiocFileCache package](https://bioconductor.org/packages/release/bioc/html/BiocFileCache.html) must be installed to make use of the `load_RDSdata_fromBox()` function. Otherwise, just download the RDS object directly and use `readRDS()`.

```{r load_SCE, echo=TRUE, message=TRUE}
fls <- fread("data/data_links.txt", header=TRUE)
sln <-  load_RDSdata_from_Box(shared_link = fls[File_name == "sce_integrated_pLNSamples_filtered.rds"]$Direct_link)
sln
```
```{r define_colors}
## define color scheme
popcols <- c("#1E498F", "#629DD4","#5E3DD8") # dark blue, light blue, intermediate blue
names(popcols) <- c("AIC","transition","AMC")

## expression gradient colors
grad_cols <- c("grey85","yellow","orange","firebrick4")
express_cols <-  c("grey85","tan1","tomato1","tomato2","firebrick3","firebrick4")

## helper function
cf <- function(cluster_as_character){
    factor(cluster_as_character, levels = c("AIC","transition","AMC"), ordered = TRUE)
}
```

>Note: In the final manuscript, 'AIC' where called 'AP' cells, 'transition cells' are 'intermediate cells' and 'AMC' are 'AM'.

## Mimicking FACS plots


```{r facs_imitation}
com <- make_long_dt(sln, features = "Tcf7",
    include_metaData = c("Tissue","Mouse","CD39","CD62L","population"), 
    exprs_values="logcounts")

ggplot(com, aes(x = CD39, y = CD62L, color = cf(population))) + 
    geom_point(shape = 1, size = 2.5, alpha = .5, stroke=1) +
    scale_color_manual(values = popcols) +
    facet_grid(~Tissue) + ggtitle("ADT: CD62L vs. CD39") +
    guides(colour = guide_legend(override.aes = list(alpha = 1, size = 4, shape = 15))) +
    theme(panel.grid = element_blank())
```

## Cells per cluster per mouse

```{r cells_per_cluster, fig.width = 14, fig.height = 10, cache=FALSE}
p1 <- colData(sln)[, c("Tissue","Mouse","population","cell")] %>% as.data.table %>% 
    ggplot(., aes(x = Tissue ,fill = cf(population))) + geom_bar(position = "fill") +
    scale_fill_manual(values = popcols,name = "pLN population") +
    theme(legend.position = "none", panel.grid = element_blank())
    
p2 <-  colData(sln)[, c("Tissue","Mouse","population","cell")] %>% as.data.table %>% 
    ggplot(., aes(x = Mouse, fill = cf(population))) + 
    geom_bar() + coord_flip() + 
    scale_fill_manual(values = popcols, name = "pLN population") + 
    theme(panel.grid = element_blank(), legend.position = "bottom")+
    ggtitle("Cells per cluster")

p3 <-  colData(sln)[, c("Tissue","Mouse","population","cell")] %>% as.data.table %>% 
    ggplot(., aes(x = Mouse, fill = cf(population))) + 
    geom_bar(position = "fill") + coord_flip() + ylab("fraction") +
    scale_fill_manual(values = popcols, name = "pLN population") + 
    theme(panel.grid = element_blank(),legend.position = "bottom")

p1 + p2 + p3 + plot_layout(widths = c(1,4,4))
```


### Diffusion map + trajectory

```{r}
## loads an object named line.data that contains the results of the TSCAN
## run to determin pseudotimes and trajectories
load_data_from_Box(fls[File_name == "tscan_lineData_DiffMap.rda"]$Direct_link, load_rds = FALSE)
```

```{r diffMap_trajectory,fig.width = 9, cache=FALSE}
plot_reducedDim_from_sce(sln, color_by = "population",
     which_reddim = "diffusionMap", 
     size_by = 2.5, alpha=.2, set_color=FALSE, 
     remove_rug=TRUE)  +  scale_color_manual(values=popcols) + 
    xlab("DC1") + ylab("DC2") +
    geom_line(data=line.data, mapping=aes(x=DC1, y=DC2, group=edge), inherit.aes = FALSE, lwd=1.5) +
    ggtitle("Diffusion map with TSCAN-predicted trajectory")
```

### Gene expression in diffusion map

```{r genes_of_interest_in_diffMap, fig.width = 12, fig.height = 15, cache=FALSE}
p1 <- plot_reducedDim_from_sce(sln, color_by = "CD62L",
     which_reddim = "diffusionMap", 
     size_by = 1, alpha=.2, set_color=FALSE, remove_rug=TRUE)  +  
    scale_color_gradientn(colors = grad_cols) + 
  #  theme(legend.position = "none") + 
    ggtitle("CD62L (ADT)") +
    xlab("DC1") + ylab("DC2")
p2 <- plot_reducedDim_from_sce(sln, color_by = "Sell",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, remove_rug=TRUE)  +  
    scale_color_gradientn(colors = express_cols) + 
     ggtitle("Sell (expression)") +
   # theme(legend.position = "none") + 
    xlab("DC1") + ylab("DC2")

p5 <- plot_reducedDim_from_sce(sln, color_by = "CD39",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, 
     remove_rug=TRUE)  +  
    scale_color_gradientn(colors = grad_cols) + 
     ggtitle("CD39 (ADT)") +
    #theme(legend.position = "none") + 
    xlab("UMAP1") + ylab("UMAP2")
p6 <- plot_reducedDim_from_sce(sln, color_by = "Entpd1",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, 
     remove_rug=TRUE)  +  
    scale_color_gradientn(colors = express_cols) + 
     ggtitle("Entpd1 (expression)") +
    #theme(legend.position = "none") + 
    xlab("DC1") + ylab("DC2") 

p7 <- plot_reducedDim_from_sce(sln, color_by = "PD-1",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, 
     remove_rug=TRUE)  +  
    scale_color_gradientn(colors = grad_cols) +
    ggtitle("PD-1 (ADT)") +
    #theme(legend.position = "none") + 
    xlab("DC1") + ylab("DC2") 
p8 <- plot_reducedDim_from_sce(sln, color_by = "Pdcd1",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, 
     add_cell_info = "Tissue", remove_rug=TRUE)  +  
    scale_color_gradientn(colors = express_cols) +
     ggtitle("Pdcd1 expression") +
    #theme(legend.position = "none") + 
   xlab("DC1") + ylab("DC2")

p4 <- plot_reducedDim_from_sce(sln, color_by = "Tcf7",
     which_reddim = "diffusionMap", exprs_values = "logcounts",
     size_by = 1, alpha=.2, set_color=FALSE, 
     remove_rug=TRUE)  +  
    scale_color_gradientn(colors = express_cols) + 
     ggtitle("Tcf7 (expression)") +
    #theme(legend.position = "none") + 
   xlab("DC1") + ylab("DC2")

(p1 + p2) / (p5 + p6) / (p7 + p8) / (p4 + p4)
```


### Genes' expression plotted along pseudotime

```{r expression_along_pseudotime, message=FALSE, fig.width = 7, fig.height = 3}
for(GN in c("Tcf7","Sell","Entpd1","Pdcd1","Lef1")){
    tt <- make_long_dt(sln, features = GN,
        exprs_values = "logcounts",
        include_metaData = c("Mouse","population","TSCANAMC"))
 P <- ggplot(tt, aes(x = TSCANAMC, y = logcounts)) +
     #ggbeeswarm::geom_quasirandom(size=.5, alpha=.2,shape=1, groupOnX = FALSE, aes(color = population)) + scale_color_manual(values = cc) +
     geom_smooth(method="auto",se=TRUE, fullrange=FALSE, level=0.95, color = "black") +
     ggtitle(GN, subtitle = "Cells sorted acc. to pseudotime") +
     facet_grid(feature_name ~ ., scales = "free_y") +
     theme(legend.position = "none", panel.grid = element_blank()) +
     xlab("pseudotime")
 print(P)
}
```


## Marker genes between clusters

### Markers for both directions

```{r marker_gene_detection, message=FALSE, cache=FALSE}
clust.markers <- scran::findMarkers(sln,
    block=sln$Sample, group = sln$population)

mks <- extract_markers(sln, 
    marker_search_result = clust.markers, FDR_thresh = 0.05,
    rank_thresh = 50, comp_name = "k40_allSampls_")
mks <- mks[!grepl("Rp[sl]", gene_symbol)]
```

```{r eval=FALSE}
clust.markers.df <- lapply(clust.markers, as.data.frame)
```

```{r mks_genes_hm, fig.height = 18, cache=FALSE}
sln$pseudotime <- sln$TSCANAMC
plot_marker_heatmap(sln, gns=mks$gene_symbol, 
    exprs_values = "logcounts",
    #cells = head(colnames(sln)),
   scale="row", n_quant_breaks = 300, col_palette = c("mediumorchid","black","yellow"),
    fontsize_row = 8,
    sort_cells_by = "pseudotime",
       add_cell_annotation = data.frame(
        row.names = colnames(sln),
        pseudotime = sln$pseudotime,
        population = sln$population,
        CD62L = sln$CD62L, CD127 = sln$CD127,
        CD38=sln$CD38, CD39 = sln$CD39, PD1 = sln$`PD-1`,
     #   CD73=sln$CD73,
        CD44 = sln$CD44),
     define_anno_cols = list(
        population = popcols,
        CD62L = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD127 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD38 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD39 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        PD1 = scales::brewer_pal("seq", "Reds")(5)[1:4],
     #   CD73 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD44 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        pseudotime = scales::viridis_pal()(5)
        )
    )
```


### Markers that go up in the respective cluster

We can focus the marker detection to those that go up in an individual cluster.

```{r marker_gene_detection_up, message=FALSE, cache=FALSE}
clust.markers.up <- scran::findMarkers(sln,direction="up",
    block=sln$Sample, group = sln$population,)

mks.up <- extract_markers(sln, 
    marker_search_result = clust.markers.up, FDR_thresh = 0.05,
    rank_thresh = 50, comp_name = "k40_allSampls_up_")
mks.up <- mks.up[!grepl("Rp[sl]", gene_symbol)]
```

```{r results='asis',eval=FALSE}
mks.up[classify == "unique"] %>% as.data.frame %>% kable(., row.names = FALSE, padding = 0, longtable = TRUE) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```
```{r mksUp_genes_hm, fig.height = 24, cache=FALSE}
plot_marker_heatmap(sln, gns=mks.up$gene_symbol, 
    exprs_values = "logcounts",
    #cells = head(colnames(sln)),
   scale="row", n_quant_breaks = 300, col_palette = c("mediumorchid","black","yellow"),
    fontsize_row = 8,
    sort_cells_by = "TSCANAMC",
       add_cell_annotation = data.frame(
        row.names = colnames(sln),
           pseudotime = sln$TSCANAMC,
        cluster = sln$population,
        CD62L = sln$CD62L,         CD127 = sln$CD127,
        CD38=sln$CD38, CD39 = sln$CD39, PD1 = sln$`PD-1`,
      #  CD73=sln$CD73,
        CD44 = sln$CD44),
     define_anno_cols = list(
        cluster = popcols,
        CD62L = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD127 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD38 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD39 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        PD1 = scales::brewer_pal("seq", "Reds")(5)[1:4],
     #   CD73 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD44 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        ps1 = scales::viridis_pal()(5), ps2 = scales::viridis_pal()(5)
        ),
    main = "Marker genes of the three populations\n(FDR 5%, Rank <= 50)"
    )
```


```{r hm_cytokines, fig.height = 8, echo=FALSE, eval=FALSE}
# Cytokines, chemokines etc
cytokines <- c("Ccr2","Cxcr3","Cxcr4","Cx3cr1","S1pr1","Itga1","Itga4","Itgae","Itgb1",'Itgb7','Cd44','Cd69',"Cxcr5","Sell","Klf2", "Il18r1","Ifngr1")

plot_marker_heatmap(sln, gns=cytokines, 
    exprs_values = "logcounts",
    fontsize_row = 12, 
    #col_palette = c("ivory",express_cols[-1]),
    scale="row", n_quant_breaks = 600, 
    col_palette = c("mediumorchid","black","yellow"),
    sort_cells_by = "TSCANAMC",
       add_cell_annotation = data.frame(
        row.names = colnames(sln),
        pseudotime = sln$TSCANAMC,
        cluster = sln$population,
         CD62L = sln$CD62L,         CD127 = sln$CD127,
        CD38=sln$CD38, CD39 = sln$CD39, PD1 = sln$`PD-1`,CD44 = sln$CD44),
     define_anno_cols = list(
        cluster = popcols,
        CD62L = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD127 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD38 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD39 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        PD1 = scales::brewer_pal("seq", "Reds")(5)[1:4],
     #   CD73 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD44 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        pseudotime = scales::viridis_pal()(5)
        ),
     main = "Selected signaling molecules"
    )
```

# Sofia's selected genes of interest

```{r }
goiDF <- read.table(file="data/fig5_genes_heatmap.csv", sep = ",", header = TRUE)

plot_marker_heatmap(sln,
    gns=goiDF$genes.for.5b.heatmap, 
    exprs_values = "logcounts",
    scale="row",
    col_palette = c(rep("mediumorchid1",7),"mediumorchid3","mediumorchid4","black","yellow3","yellow2",rep("yellow1",7)),
    fontsize_row = 8,
    sort_cells_by = "TSCANAMC",
    add_cell_annotation = data.frame(
        row.names = colnames(sln),
        pseudotime = sln$TSCANAMC,
        cluster = sln$population,
        CD62L = sln$CD62L, CD127 = sln$CD127,
        CD38=sln$CD38, CD39 = sln$CD39, PD1 = sln$`PD-1`,
        CD44 = sln$CD44),
     define_anno_cols = list(
        cluster = popcols,
        CD62L = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD127 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD38 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD39 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        PD1 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        CD44 = scales::brewer_pal("seq", "Reds")(5)[1:4],
        ps1 = scales::viridis_pal()(5), ps2 = scales::viridis_pal()(5)
        ),
    main = "Selected marker genes"
    )
```


# GO term enrichments

Ignoring ribosomal genes.

```{r message=FALSE, warning=FALSE}
## get markers
cm.up <- scran::findMarkers(sln,
    block=sln$Sample, group = sln$population, direction = "up")
mks.up <- extract_markers(sln, 
    marker_search_result = cm.up, FDR_thresh = 0.05,
    rank_thresh = 50)
mks.up <- mks.up[ !grepl("^Rp[ls]", gene_symbol)] # removing ribosomal genes

## get ENTREZ IDs
eg <-  clusterProfiler::bitr(
    unique(mks.up$gene_symbol), 
    fromType="SYMBOL", toType="ENTREZID",
    OrgDb="org.Mm.eg.db") %>%  as.data.table
setnames(eg, names(eg), c("gene_symbol", "entrez"))
eg <- mks.up[eg, on = "gene_symbol"]

## generate list of ENTREZ IDs of interest
clstcomp.list <- lapply(cm.up, function(x){
    eg[ gene_symbol %in% rownames(subset(x, FDR <= 0.05))]$entrez}
    )

## universes
all.entrez <-  clusterProfiler::bitr(rownames(sln), 
    fromType="SYMBOL", toType="ENTREZID",
    OrgDb="org.Mm.eg.db") %>%  as.data.table
all.entrez_noRibo <- all.entrez[!grepl("^Rp[ls]", SYMBOL)]

## for visualization purposes
logFClist <- lapply(cm.up, function(x){
    out <- x[, "summary.logFC"]
    names(out) <- row.names(x)
    return(out)
})
```

```{r}
load_data_from_Box(fls[object == "clstProfiler_BP"]$link, load_rds = FALSE) # loads 3 objects of cluster profiler results, one for each population (AIC/AMC/Transition)
load_data_from_Box(fls[object == "clstProfiler_MF"]$link, load_rds = FALSE)
```


## GO terms (BP) {.tabset}

Here, we first directly compare the GO term enrichments for the different clusters.
Each tab then contains the results of the individual analyses, i.e. one per cell population (AIC, transition, AMC).

If a tab does not contain an extra plot, no enrichments were found for that cell population.

### Comparison

```{r gobp_cluster_comparison, fig.width = 7, message=FALSE, warning=FALSE}
dotplot(cc.gobp) + ggtitle("Overrepresented GO BP terms") + 
     scale_color_gradientn(colours = rev(c("azure2", "darkseagreen1",
         "seagreen2","seagreen3")),  limits = c(0, 0.05))
```

- geneRatio should be number of genes that overlap gene set divided by size of gene set

### AIC

```{r gobp_aic_cnetplot, fig.height = 12, fig.width = 15}
P <- try(cnetplot(gobp_aic, showCategory = 8,
    foldChange = logFClist$AIC, 
    colorEdge = TRUE), TRUE) 
if(length(P) > 1){ 
    P + ggtitle("GO BP PWs (AIC)")
}
```


### Transition

```{r gobp_transition_cnet, fig.width = 15, fig.height = 10}
#gobp_trans <- enrichGO(gene = clstcomp.list$transition, OrgDb = org.Mm.eg.db,
#    ont = "BP", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez_noRibo$ENTREZID)
P <- try(cnetplot(gobp_trans, showCategory = 8,
    foldChange = logFClist$transition,
    colorEdge = TRUE), TRUE)  
if(length(P) > 1){ P + ggtitle("GO BP PWs (Transition population)") +  
        scale_color_gradientn(colours =c("lightsalmon", "tomato", "firebrick4"))}
```

* grey bubbles = name of the gene set at hand (e.g. "translational elongation"
    - size = number of marker genes that are part of the gene set
* edges = colored by gene set
* nodes belonging to each gene set represent individual genes (e.g. Rack1)
    - their color represents the logFC of that gene in the respective cluster compared to the expression of the gene in all other clusters


### AMC

```{r gobp_amc_cnet, fig.width = 16, fig.height = 12}
#gobp_amc <- enrichGO(gene = clstcomp.list$AMC,OrgDb = org.Mm.eg.db,
#    ont = "BP", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez_noRibo$ENTREZID)

P <- try(cnetplot(gobp_amc, showCategory = 8, foldChange = logFClist$AMC,
    colorEdge = TRUE), TRUE) 
if(length(P) > 1){ P + ggtitle("GO BP PWs (AMC)")  +  
        scale_color_gradientn(colours =c("lightsalmon", "tomato", "firebrick4"))}
```


## GO terms (MF) {.tabset}

If a tab does not contain an extra plot, no enrichments were found for that cell population.

### Comparison

```{r gomf_cluster_comparison, fig.width = 7, message=FALSE, warning=FALSE}
## visualization
dotplot(cc.gomf) + ggtitle("Overrepresented GO MF terms") + 
     scale_color_gradientn(colours =rev(c("azure2", "darkseagreen1", "seagreen2","seagreen3")), limits = c(0, 0.05))
```

- geneRatio should be number of genes that overlap gene set divided by size of gene set

### AIC

```{r gomf_aic_cnetplot, fig.height = 12, fig.width = 15}
#gomf_aic <- enrichGO(gene = clstcomp.list$AIC, OrgDb = org.Mm.eg.db,
#    ont = "MF", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez_noRibo$ENTREZID)

P <- try(cnetplot(gomf_aic, showCategory = 8, foldChange = logFClist$AIC, 
    colorEdge = TRUE), TRUE) 
if(length(P) > 1){ 
    P + ggtitle("GO MF PWs (AIC)")
}
```


### Transition

```{r gomf_transition_cnet, fig.width = 15, fig.height = 10}
#gomf_trans <- enrichGO(gene = clstcomp.list$transition, OrgDb = org.Mm.eg.db,
#    ont = "MF", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez_noRibo$ENTREZID)

P <- try(cnetplot(gomf_trans, showCategory = 8,
    foldChange = logFClist$transition,
    colorEdge = TRUE), TRUE)  
if(length(P) > 1){ P + ggtitle("GO MF PWs (Transition population)") +  
        scale_color_gradientn(colours =c("lightsalmon", "tomato", "firebrick4"))}
```

* grey bubbles = name of the gene set at hand (e.g. "translational elongation"
    - size = number of marker genes that are part of the gene set
* edges = colored by gene set
* nodes belonging to each gene set represent individual genes (e.g. Rack1)
    - their color represents the logFC of that gene in the respective cluster compared to the expression of the gene in all other clusters


### AMC

```{r gomf_amc_cnet, fig.width = 16, fig.height = 12}
#gomf_amc <- enrichGO(gene = clstcomp.list$AMC,OrgDb = org.Mm.eg.db,
#    ont = "MF", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez_noRibo$ENTREZID)
P <- try(cnetplot(gomf_amc, showCategory = 8, foldChange = logFClist$AMC,
    colorEdge = TRUE), TRUE) 
if(length(P) > 1){ P + ggtitle("GO MF PWs (AMC)")  +  
        scale_color_gradientn(colours =c("lightsalmon", "tomato", "firebrick4"))}
```

# SessionInfo

```{r}
sessionInfo()
```

