---
title: "Combining our data with Schauder's Tcm and Yao's progLikes and MPECs"
author: "Friederike Dündar | ABC @ WCM"
date: "August 2021"
output:
    html_document:
        code_folding: hide
        toc: true
        toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r cache=FALSE, message=FALSE}
library(data.table); library(magrittr)
library(ggplot2); theme_set(theme_bw(base_size = 14))
library(patchwork)
library(SingleCellExperiment)
library(pcaExplorer); library(org.Mm.eg.db)
source("src/marker_gene_extract_and_plot.R")
source("src/load_data_from_box.R")
library(pheatmap)
library(org.Mm.eg.db); library(clusterProfiler)
```

## Integrating data sets

* Yao et al (2019): day 7 CD8 T cells from acute (`Arm`) and chronic (`Cl13`) infection
    - will focus on the memory precursors and progenitor-like cells (which are from repl. 1 only)
* Schauder et al (2021): day 129 following acute infection

For details of how we retrieved the public data sets, see `Schauder2021.Rmd` and `Yao2019.Rmd`.

```{r integration, eval=FALSE}
## Code here does not need to be executed, it's simply for documentation purposes
## to show the details of how we generated the composite integrated file that
## we're providing for download
library(magrittr)
library(batchelor)
library(scran)
library(scater)
library(patchwork)

fnall <- "sce_integrated_with_YaoProgsMpecs-and-Schauder.rds"
fnmerged <- "sceMultiout_integrated_with_YaoProgsMpecs-and-Schauder.rds"

## MAKE A LIST of INDIVIDUAL SCEs ==============================================
## our data -------------
sln <- load_data_from_Box("https://wcm.box.com/shared/static/2wvbdfs3vja2cnlckcqpk20eu1693o8o.rds") 
rownames(sln) <- rowData(sln)$ID
scel <- lapply(unique(sln$Sample), function(x){
    out.sce <- sln[, sln$Sample == x]
    assay(out.sce, "logcounts") <- NULL
    colData(out.sce) <- colData(out.sce)[, c("Sample","Barcode","cell")]
    rownames(out.sce)  <- rowData(out.sce)$ID
    return(out.sce)
})
names(scel) <- unique(sln$Sample)

## Schauder -------------
scs <- readRDS("../2021-08_Schauder2021/sce_integrated_with_Schauder2021.rds")
assay(scs, "logcounts") <- NULL
scs <- scs[,!grepl("^pLN",scs$Sample)]
colData(scs) <- colData(scs)[, c("Sample","Barcode","cell")]
rownames(scs)  <- rowData(scs)$ID
scel$Schauder <- scs

## Yao -------------
scy <- readRDS("../2021-08_Yao2019/sce_integrated_with_Yao2019.rds")
assay(scy, "logcounts") <- NULL
scy <- scy[,!grepl("^pLN",scy$Sample)]

yaolabs <- read.table("../2021-08_Yao2019/Yao2019_metadata.txt", header = FALSE, skip = 1)
names(yaolabs) <-  c("YaoCell","nGene","nUMI","orig.ident","percent.mito","res.0.5","res.1","res.1.5")
## res.0.5 = clusters; 10 = progs, 7 = MPecs
kp_cells <- subset(yaolabs, res.0.5 %in% c(10, 7)) %>% .$YaoCell
scy$YaoCell <- paste(gsub("_[12]$", "", scy$Sample), gsub("-[0-9]$","", scy$Barcode), sep = "_")

for(i in unique(scy$Sample)){
    out.sce <- scy[, scy$Sample == i]
    out.sce <- out.sce[, out.sce$YaoCell %in% kp_cells]
    colData(out.sce) <- colData(out.sce)[, c("Sample","Barcode","cell")]
    rownames(out.sce)  <- rowData(out.sce)$ID
    scel[[i]] <- out.sce}

## PREPPING ============================================================
rd_qc <- lapply(scel,perFeatureQCMetrics)
for(x in seq_along(scel)){rowData(scel[[x]])$qc.mean <- rd_qc[[x]]$mean}
scel2 <- lapply(scel, function(x){ x[rowData(x)$qc.mean > 0.001,]})
## combine
universe <- Reduce(intersect, lapply(scel2, rownames))
scel2 <- lapply(scel2, "[", i=universe)
# generate logcounts
normed.sce <- do.call(multiBatchNorm, scel2) # returns a list
# Identifying a set of HVGs using stats from all batches, using logcounts
all.dec <- lapply(normed.sce, modelGeneVar)
combined.dec <- do.call(combineVar, all.dec)
combined.hvg <- getTopHVGs(combined.dec, n=2000)

## MERGING with MNN ====================================================
## prep
combined <- noCorrect(normed.sce)
assayNames(combined) <- "logcounts"
combined$Sample <- combined$batch
combined$batch <- NULL
set.seed(1010100)
## progressively merge cells from each sample in each batch until all cells 
## are mapped onto a common coordinate space
multiout <- fastMNN(combined, batch=combined$Sample, subset.row=combined.hvg)
# Renaming metadata fields for easier communication later.
multiout$Sample <- multiout$batch

## UMAP----------------------------------
set.seed(10101010)
multiout <- runUMAP(multiout, dimred="corrected")

## Clustering -----------------------------
g <- buildSNNGraph(multiout, use.dimred="corrected", k = 20)
clusters <- igraph::cluster_louvain(g)
multiout$cluster_with_SchauderYao_k20 <- factor(clusters$membership)

g <- buildSNNGraph(multiout, use.dimred="corrected", k = 50)
clusters <- igraph::cluster_louvain(g)
multiout$cluster_with_SchauderYao_k50 <- factor(clusters$membership)

g <- buildSNNGraph(multiout, use.dimred="corrected", k = 10)
clusters <- igraph::cluster_louvain(g)
multiout$cluster_with_SchauderYao_k10 <- factor(clusters$membership)

g <- buildSNNGraph(multiout, use.dimred="corrected", k = 5)
clusters <- igraph::cluster_louvain(g)
multiout$cluster_with_SchauderYao_k5 <- factor(clusters$membership)

saveRDS(multiout,file = fnmerged)

# generate composite file for the combined file of all shared genes ============
## combine
universe <- Reduce(intersect, lapply(scel, rownames))
scel <- lapply(scel, "[", i=universe)
comb.mat <- lapply(scel, function(x) counts(x)) %>% do.call(cbind, .)
colnames(comb.mat) <- unlist(lapply(scel, function(x) colnames(x)))

### rowData
rd <- rowData(scel[[1]])[, c("ID","Symbol")]
rd <- rd[rownames(comb.mat),]

## colData 
cd <- lapply(scel, function(x) colData(x)[, c("Sample","Barcode","cell")]) %>% do.call(rbind, .)
cd <- cd[colnames(comb.mat),]

scAll <- SingleCellExperiment(
    assays = list(counts = comb.mat), 
    colData = cd, rowData = rd)

## add redDims from the merged data set
rdu <- reducedDim(multiout, "UMAP") 
reducedDim(scAll, "UMAP") <- rdu[colnames(scAll),]
reducedDim(scAll, "PCA_corr") <- reducedDim(multiout, "corrected")

## add log-counts
qckclst <- quickCluster(scAll, method = "igraph", min.mean = 0.1)
scAll <- computeSumFactors(scAll, min.mean=0.1, cluster = qckclst)
scAll <- scater::logNormCounts(scAll)

saveRDS(scAll, file = fnall)
colData(multiout) %>% saveRDS("colData_multiout_YaoProgsMpecs-and-Schauder.rds")

## fix ColData ======================================================
cd2 <- colData(multiout)
cd2 <- as.data.frame(cd2)
cd2$cell <- rownames(cd2)
cd2 <- cd2[, c("cell",grep("cluster", names(cd2), value=TRUE))]
newcd <- merge(colData(scAll), cd2, by = "cell") 

## add YAO LABELS -----------------------------------------------
newcd$YaoCell <- paste(gsub("_[12]$", "", newcd$Sample), gsub("-[0-9]$","", newcd$Barcode), sep = "_")
newcd <- merge(newcd, yaolabs, by="YaoCell", all.x = TRUE)

## add OUR LABELS ------------------------------------------------
cdus <- colData(sln)
newcd <- cdus[, c("cell","CD62L","label")] %>% as.data.frame %>% 
    merge(newcd, ., on = "cell", all.x = TRUE)
newcd$label <- ifelse(!is.na(newcd$label), newcd$label,
    ifelse(grepl("GSM3732587", newcd$Sample), "Schauder",
        ifelse(grepl("^D", newcd$Sample),  paste0("Yao_", gsub("_1$", "", gsub("_P14","",newcd$Sample))),
            "no.clue")))
## 10 = progs, 7 = MPecs
newcd$label2 <- ifelse(is.na(newcd$res.0.5), newcd$label,
    ifelse(newcd$res.0.5 == 10, "ProgLike", 
        ifelse(newcd$res.0.5 == 7, paste(newcd$label, "MPECS", sep = "_"), "no.clue")))
newcd$label2 <- ifelse(newcd$label2 == "Schauder", "Tcm", newcd$label2)

#> table(newcd$label2)
#             AIC               AMC          ProgLike               Tcm 
#             1704              1335               114               535 
#       transition  Yao_D7_Arm_MPECS Yao_D7_Cl13_MPECS 
#             2016               386                 2 


## finalize colData
rownames(newcd) <- newcd$cell
colData(scAll) <- newcd[colnames(scAll),]

## adjust the naming scheme for our populations
scAll$label <- ifelse(scAll$label == "AIC","AP", ifelse(scAll$label == "transition", "intermediate", ifelse(scAll$label == "AMC", "AM", scAll$label)))
scAll$label2 <- ifelse(scAll$label2 == "AIC","AP", ifelse(scAll$label2 == "transition", "intermediate", ifelse(scAll$label2 == "AMC", "AM", scAll$label2)))
## remove the lonely two cells classified as D7-Cl13-MPECS
scAll <- scAll[, scAll$label2 != "Yao_D7_Cl13_MPECS"]
scAll$label2 <- gsub("Yao_D7_Arm_", "", scAll$label2)
scAll$label2 <- factor(scAll$label2, levels = c("AP","intermediate","AM","MPECS","ProgLike", "Tcm"), ordered = TRUE)
scAll$Ref <- ifelse(grepl("Yao",scAll$label), "Yao", ifelse(grepl("Schauder", scAll$label), "Schauder", "Gearty"))


rownames(scAll) <- scater::uniquifyFeatureNames(ID = rowData(scAll)$ID, 
    names = rowData(scAll)$Symbol)

saveRDS(scAll, file = fnall)
```

```{r load_data}
sc3 <- load_RDSdata_from_Box(shared_link = "https://wcm.box.com/shared/static/ixtkzzow5b4r2ck3u6zit2jtep7dunzv.rds")
```

```{r define_colors}
## define color schemes ---------------------------------------------------------
popcols <- c("#1E498F", "#629DD4","#5E3DD8", # dark blue, light blue, intermediate blue
    "#228B22", #"bisque1",
    "#9ED9B9",#"lightpink1", 
    "maroon1") 
names(popcols) <- c("AP","intermediate","AM", 
    "Tcm",
    "MPECS","ProgLike")

sample_cols <- c(
    "D7_P14_Cl13_1" = "sienna4", "D7_P14_Arm_1"="sienna2",
    "GSM3732587_Day129"="wheat3",
    "pLN_1"="blue","pLN_2"="dodgerblue1","pLN_3"="lightskyblue")


cluster_cols <- c('#B1E2F9','limegreen','grey30','#FFC914','#0066E2','#FC71E9','#00E5CA','#E31A1C','grey83','#FF7F00','#378C07','#6A3D9A','#FFFF99','#B15928')
names(cluster_cols) <- sort(unique(sc3$cluster_with_SchauderYao_k20))
cluster_cols <- cluster_cols[sort(unique(sc3$cluster_with_SchauderYao_k20))]
```

## Global comparisons 

We can pretend that the individual cell types of interest are pseudo-bulk samples,
i.e. we can sum up the reads across all cells of the same label-sample combination
(e.g. all cells from Sample "pLN_1" that are labelled as "AP" will form one pseudo-bulk
sample "pLN_1_AP").
This helps us to apply more robust bulk-RNA-seq methods for global comparisons across the samples.

```{r message=FALSE}
summed.scf <- scater::aggregateAcrossCells(sc3,
    ids=colData(sc3)[,c("label2","Sample")])
library(edgeR)
y <- DGEList(counts(summed.scf), samples=colData(summed.scf))
keep <- filterByExpr(y, group=summed.scf$Sample)
y <- y[keep,]
y <- calcNormFactors(y)
```

### PCA

```{r PCA, fig.width = 15, fig.height = 8}
PCA <- prcomp(t(cpm(y, log = TRUE)), scale = F)
percentVar <- round(100*PCA$sdev^2/sum(PCA$sdev^2),1)
dataGG = data.frame(PC1 = PCA$x[,1],PC2 = PCA$x[,2], 
    PC3 = PCA$x[,3],PC4 = PCA$x[,4],
    sampleName = row.names(y$samples),
    y$samples)
dataGG$MySampleName <- paste(dataGG$Sample.1, dataGG$label2.1, sep = "_")

p1 <- ggplot(dataGG, aes(x = PC1, y = PC2, fill = label2.1, shape = Ref)) +
    geom_point(size = 8,  alpha = .85) +
    xlab(paste0("PC1, VarExp:", round(percentVar[1],4))) +
    ylab(paste0("PC2, VarExp:", round(percentVar[2],4))) +
    scale_fill_manual(values = popcols, name = "") + 
    scale_shape_manual(values = c(21,22,23), name = "") +
   theme(legend.position = "bottom")

p2 <- ggplot(dataGG, aes(x = PC1, y = PC2, color = label2.1, label = MySampleName)) +
    geom_text(size = 6) + scale_color_manual(values = popcols) +
    theme(legend.position = "none")

p1 | p2
```

```{r pca_loadings, message=FALSE, warning=FALSE, eval=TRUE, fig.width=10, fig.height=6.5, context="data"}
par(mfrow=c(2,1))
hi_loadings(PCA,whichpc = as.integer(1),topN = 20)
hi_loadings(PCA,whichpc = as.integer(2),topN = 20)
```


```{r hclusting, fig.height=8, fig.width = 10}
hc <- hclust( as.dist(1-cor(cpm(y, log = TRUE))), method="complete")
hc$labels <- dataGG$MySampleName
#plot(hc)
```

```{r hclust_mod, fig.height = 4.5, fig.width =7, message=FALSE}
library(dendextend)
dend <- as.dendrogram(hc)
labels_colors(dend) <- c(rep(popcols[["AP"]], 3), popcols[["AM"]], rep(popcols[["intermediate"]], 3), rep(popcols[["AM"]], 2), popcols[["Tcm"]], popcols[["MPECS"]], popcols[["ProgLike"]])

par(mar=c(15,4,4,2))
rotate(dend, c( 'pLN_1_AP', 'pLN_3_AP', 'pLN_4_AP', 'pLN_1_intermediate', 'pLN_3_intermediate', 'pLN_4_intermediate', 'pLN_1_AM', 'pLN_3_AM', 'pLN_4_AM', 'D7_P14_Arm_1_MPECS', 'D7_P14_Cl13_1_ProgLike','GSM3732587_Day129_Tcm')) %>% plot
```


## Single-cell level: markers distinguishing the populations

>comparing AP vs ProgLike vs Tcm and so on

Taking all genes that are either up- or down in any one of the populations (FDR <= 1%) and that are at least in the top 50 for each population.

```{r marker_gene_detection_pops, message=FALSE, cache=FALSE}
LabelMarkers <- scran::findMarkers(sc3,
    group = sc3$label2)

goi <- lapply(LabelMarkers, function(x) rownames(subset(as.data.frame(x), FDR <= 0.01 & Top <= 50))) %>% unlist %>% unique
```

```{r fig.height = 20, fig.width = 12}
## with Tcf7 as a color bar 
plot_marker_heatmap(sc3, 
    gns=unique(c("Tox",goi)), 
    gns_as_cellAnnotation = c("Tcf7","Tox"), exclude_genes = "Tox",
    exprs_values = "logcounts",
    show_HKgenes = "both",
    n_quant_breaks = 300,
   scale="row",col_palette = c("mediumorchid","black","yellow"),
    fontsize_row = 7,
  add_cell_annotation = data.frame(
        row.names = colnames(sc3),
        #cluster = sc3$cluster_with_SchauderYao_k20,
        population = sc3$label2),
    define_anno_cols = list(
        #cluster = cluster_cols,
        population = popcols,
        Tox = scales::brewer_pal("seq", "Reds")(5)[1:4],
        Tcf7 = scales::brewer_pal("seq", "Reds")(5)[1:4]),
    main = "Marker genes of the individual populations\n(FDR 1%, Rank <= 50)"
    )
```

### GO term enrichments

Determining the top enriched GO terms for markers that are specifically overexpressed in the different populations.

```{r message=FALSE, warning=FALSE, eval=FALSE}
## marker genes
cm.up <- scran::findMarkers(sc3,
    group = sc3$label2, direction = "up")
mks.up <- extract_markers(sc3, 
    marker_search_result = cm.up, FDR_thresh = 0.01,
    rank_thresh = 300)

## get ENTREZ IDs
all.entrez <-  clusterProfiler::bitr(rownames(sc3), 
    fromType="SYMBOL", toType="ENTREZID",
    OrgDb="org.Mm.eg.db") %>%  as.data.table
setnames(all.entrez, names(all.entrez), c("gene_symbol", "entrez"))

## generate list of ENTREZ IDs of interest
eg <- all.entrez[mks.up, on = "gene_symbol"]
## focus on those that are UP and UNIQUELY so per population
clstcomp.list <- lapply(names(cm.up), function(x) eg[up_in == x & classify == "unique"]$entrez )
names(clstcomp.list) <- names(cm.up)
```

```{r gobp_cluster_comparison, eval=FALSE}
cc.gobp <- compareCluster(clstcomp.list, fun = "enrichGO", OrgDb = org.Mm.eg.db,
    ont = "BP", pvalueCutoff=0.05, readable=TRUE, universe = all.entrez$entrez)
```
```{r include=FALSE}
load("YaoSchauderGearty_markerGenes_GOBP.rda")
```

```{r fig.width = 10, fig.height = 7, message=FALSE, warning=FALSE}
## visualization
dotplot(cc.gobp, showCategory = 10) +
    ggtitle("Overrepresented GO BP terms for markers that are\nuniquely upregulated in either population") + 
     scale_color_gradientn(colours =rev(c("azure2", "darkseagreen1", "seagreen2","forestgreen")), limits = c(0, 0.05))
```

- geneRatio should be number of genes that overlap gene set divided by size of gene set

Here are some of the top enriched GO terms in detail:

```{r fig.width = 15, fig.height = 15}
cnetplot(cc.gobp) + scale_fill_manual(values = popcols, name = "")
```

## Session Info

```{r}
sessionInfo()
```

