---
title: "V(D)J-seq based plots for Gearty et al. (2021)"
author: "Friederike Dündar | Weill Cornell Medicine"
date: "2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '6'
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    toc_depth: 6
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, cache=FALSE}
library(SingleCellExperiment)
library(magrittr); library(data.table)
library(ggplot2); theme_set(theme_bw(base_size = 14))
library(patchwork)
library(ggalluvial)
library(scales)
source("src/utils_SCE.R")
source("src/load_data_from_box.R")

tf <- function(tissue_as_character){
    factor(tissue_as_character, levels = c("pLN.Tcf1hi","pLN.Tcf1lo","pancreas"), ordered = TRUE)
}

popf <- function(population_as_character){
    factor(population_as_character, 
        levels = c("pancreas","pLN.Tcf1hi,pancreas","pLN.Tcf1hi,pLN.Tcf1lo,pancreas",
            "pLN.Tcf1lo,pancreas","pLN.Tcf1hi,pLN.Tcf1lo","pLN.Tcf1lo","pLN.Tcf1hi"), 
        ordered = TRUE)
}
```

The plots here are based on the results of a V(D)J-seq experiment where CD8 T cells from pancreatic lymphnodes (pLN) and the pancreas (panc) were obtained and their TCR gene sequences were profiled with 10X Genomics' V(D)J sequencing strategy.


```{r define_colors}
## define color scheme
tcfcols <- c("#1E498F", "#5E3DD8","#D72027")
names(tcfcols) <- c("pLN.Tcf1hi","pLN.Tcf1lo","pancreas")

shared_cols <- c("khaki1","#39568CFF")

## expression gradient colors
grad_cols <- c("grey85","yellow","orange","firebrick4")
express_cols <-  c("grey85","tan1","tomato1","tomato2","firebrick3","firebrick4")
```


```{r load_tcr_data}
fls <- fread("data/data_links.txt", header=TRUE)
tcrCells <- load_RDSdata_from_Box(fls[object=="vdj"]$link)

## FILTERING =================================================================
## keep those where we have info from the same mouse for both panc & pLN,
## as well as max. one TRA and/or one TRB sequence
tcrCells <- tcrCells[pairedExp == TRUE & chains %in% c("TRA","TRB","TRA;TRB")]
tcrCells[Tissue == "panc", population := "pancreas"]
## remove those for which we don't have scRNA-seq info from the pLN, 
## i.e. their population info is missing
tcrCells[is.na(population), population := "pLN"]; tcrCells <- tcrCells[population != "pLN"] 
## define Tcf status
tcrCells[ , Tcf.status :=
        ifelse(population %in% c("AIC","transition"), "pLN.Tcf1hi",
            ifelse(population=="AMC", "pLN.Tcf1lo", "pancreas"))]

## count ======================================================================
clono_per_pop <-  tcrCells[, .N, by = c("new_cid","Tcf.status", "Mouse")]
shared_across_pops <- clono_per_pop[, .N, new_cid] %>% .[N>1] %>% .$new_cid
clono_per_pop[, shared_between_pops := ifelse(new_cid %in% shared_across_pops, TRUE, FALSE)]
clono_per_pop[, pseudo_clono.ID := ifelse(shared_between_pops, paste0(new_cid, Mouse), "unique")]
clono_per_pop[, allPops := paste(sort(unique(Tcf.status)), collapse=","), by = new_cid]
## fix the order of the labels
clono_per_pop[ allPops=="pancreas,pLN.Tcf1hi", allPops := "pLN.Tcf1hi,pancreas"]
clono_per_pop[ allPops=="pancreas,pLN.Tcf1hi,pLN.Tcf1lo", allPops := "pLN.Tcf1hi,pLN.Tcf1lo,pancreas"]
clono_per_pop[ allPops=="pancreas,pLN.Tcf1lo", allPops := "pLN.Tcf1lo,pancreas"]
```

### Clonotype frequencies

```{r fig.width = 15, fig.height = 8, message=FALSE}
p1 <-  ggplot(clono_per_pop, aes(y = shared_between_pops, x = log2(N), fill = shared_between_pops)) +
    ggridges::geom_density_ridges2(scale=0.95) +
     scale_fill_manual(values=shared_cols) +
     ylab("number of cells") +   xlab("log2(number of cells)") +
    facet_wrap(~tf(Tcf.status))  +
    theme(legend.position = "bottom", panel.grid = element_blank())
p2 <- ggplot(clono_per_pop, aes(x = shared_between_pops, y = N, fill = shared_between_pops)) +
    geom_boxplot(fill = "white", notch=TRUE) + 
    ggbeeswarm::geom_quasirandom(size=2, shape=21, alpha=.7) + 
    facet_wrap(~tf(Tcf.status)) +
    scale_fill_manual(values=shared_cols) +
    xlab("clonotype found in more than one cell population\n(from the same mouse)") +
     ylab("number of cells") +
    theme(legend.position = "none",
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_blank()) +
    scale_y_continuous(trans=log2_trans())
pw <- p1 + p2
pw + plot_annotation(title = "Number of cells per clonotype per population", 
    subtitle = "Split by whether a clonotype is unique or found across multiple populations")
```

These plots demonstrate that clonotypes that are found in just one population ("FALSE"; yellow) tend to be found in single cells (N=1) whereas clonotypes that are found across multiple populations ("TRUE"; darkblue) are usually found in multiple cells per population.

## Counting clonotypes

```{r}
clono_per_allPops <- clono_per_pop[, -c("N","Tcf.status")] %>% unique
```


```{r}
ggplot(clono_per_allPops, aes(x = popf(allPops))) + 
    geom_bar() + coord_flip() +
    theme(panel.grid = element_blank()) +
    xlab("") + ylab("number of clonotypes")
```

Here are the numbers:

```{r}
clono_per_allPops[, .N, allPops]
```


```{r fig.width = 8}
ggplot(clono_per_pop, aes(x = tf(Tcf.status), fill = shared_between_pops)) + 
    geom_bar(position = position_dodge(), color = "grey75") + 
    scale_fill_manual(values = shared_cols) + xlab("") + ylab("Number of clonotypes") +
    theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
```

And here's the same figure, just stacked:

```{r fig.height = 3}
ggplot(clono_per_pop[Tcf.status != "pancreas"], aes(x = tf(Tcf.status),
    fill = factor(shared_between_pops, levels = c(TRUE, FALSE), ordered = TRUE))) + 
    geom_bar( color = "grey75") + 
    scale_fill_manual(values = rev(shared_cols), name = "shared") + xlab("") + ylab("Number of clonotypes") +
    theme(panel.grid = element_blank()) +
    coord_flip()
```

The plots above don't show the counts at the individual mouse level, which may help simplify things.
The actual numbers underlying the plots above:

```{r }
clono_per_pop[, .N, by = c("Tcf.status","shared_between_pops")] %>% 
    dcast(Tcf.status ~ shared_between_pops, value.var = "N")
```

#### Verbose clonotype number statements

* There are `r length(unique(tcrCells$new_cid))` clonotypes **across all cells** of which `r length(shared_across_pops)` are found in more than one population ("shared").
* There are `r length(unique(tcrCells[Tissue == "panc"]$new_cid))` clonotypes in the **pancreas** of which `r length(unique(tcrCells[Tissue == "panc" & new_cid %in% shared_across_pops]$new_cid))` are shared with at least one pLN population.
    - Of these `r nrow(clono_per_allPops[grepl("pancreas", allPops)])` shared pancreas clonotypes `r nrow(clono_per_allPops[grepl("pancreas", allPops) & grepl("Tcf1hi", allPops)])` are found in the Tcf1hi population (and `r nrow(clono_per_allPops[grepl("pancreas", allPops) & grepl("Tcf1lo", allPops)])` in the Tcf1-lo population)
* There are `r length(unique(tcrCells[Tcf.status == "pLN.Tcf1hi"]$new_cid))` in the **pLN.Tcf1hi population** of which `r length(unique(tcrCells[Tcf.status == "pLN.Tcf1hi" & new_cid %in% shared_across_pops]$new_cid))` are shared with other populations.
    - Of these `r length(unique(tcrCells[Tcf.status == "pLN.Tcf1hi" & new_cid %in% shared_across_pops]$new_cid))` shared clonotypes, `length(unique(tcrCells[Tcf.status == "pLN.Tcf1hi" & new_cid %in% shared_across_pops & new_cid %in% tcfHI_and_panc]$new_cid))` are found in the pancreas, too.

```{r fig.height = 7, fig.width = 7}
gplots::venn(list(pancreas = unique(clono_per_allPops[grepl("pancreas", allPops)]$new_cid),
    pLN.Tcf1hi = unique(clono_per_allPops[grepl("Tcf1hi", allPops)]$new_cid), 
    pLN.Tcf1lo = unique(clono_per_allPops[grepl("Tcf1lo", allPops)]$new_cid)))
```


#### Per mouse

To bring the individual mouse information back in, we can use scatter plots again:

```{r}
clono_per_pop[, .N, by = c("Tcf.status","shared_between_pops","Mouse")] %>%
    ggplot(., aes(x = tf(Tcf.status), y = N, fill = shared_between_pops,label = Mouse)) + 
    ggbeeswarm::geom_quasirandom(dodge.width = .8, shape=21, groupOnX = TRUE) +
    scale_fill_manual(values = shared_cols) +
    ylab("number of clonotypes") + xlab("") +
    ggtitle("Number of clonotypes per mouse") +
    theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank()) 
```

```{r}
clono_per_pop[, .N, by = c("Tcf.status","shared_between_pops","Mouse")] %>%
    dcast(Mouse + Tcf.status ~ shared_between_pops, value.var = "N")
```

And here are the mean values:

```{r}
clono_per_pop[, .N, by = c("Tcf.status","shared_between_pops","Mouse")] %>% .[ , mean(N), by = c("Tcf.status","shared_between_pops")] %>% dcast(Tcf.status~shared_between_pops, value.var = "V1")
```

I.e., for pLN.Tcf1hi there are, on average, 54 clonotypes per mouse that are not shared (=unique), whereas we detect only 1 to 2 unique clonotypes per mouse in the pLN.Tcf1lo population.

These are the values if we do not split by shared or not-shared:

```{r}
clono_per_pop[, .N, by = c("Tcf.status","Mouse")] %>% .[ , mean(N), by = c("Tcf.status")]
```

##### Focus on unique = non-shared) ones

```{r fig.height = 7, fig.width = 13}
p1 <- clono_per_pop[shared_between_pops == FALSE] %>%
    ggplot(., aes(x = Mouse,
        fill = factor(Tcf.status, levels = c("pancreas","pLN.Tcf1lo","pLN.Tcf1hi"), ordered = TRUE))) + geom_bar() + coord_flip() +
    scale_fill_manual(values = tcfcols) +
    theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(), legend.title = element_blank()) +
    ggtitle("Number of unique clonotypes", subtitle = "clones found in just one population")

p2 <- clono_per_pop[shared_between_pops == FALSE] %>%
    ggplot(., aes(x = Mouse,
        fill = factor(Tcf.status, levels = c("pancreas","pLN.Tcf1lo","pLN.Tcf1hi"), ordered = TRUE))) + geom_bar(position = "fill") + coord_flip() +
    scale_fill_manual(values = tcfcols) +
    theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(), legend.title = element_blank()) +
    ggtitle("Number of unique clonotypes", subtitle = "clones found in just one population") + ylab("number of unique clonotypes (fraction per mouse)")

p1 + p2
```


##### Fractions

```{r fig.height = 7, fig.width = 8}
ggplot(clono_per_pop, aes(x = Mouse, fill = shared_between_pops)) + geom_bar(position = "fill") + coord_flip() +
    scale_fill_manual(values = shared_cols) +
    theme(panel.grid.minor = element_blank(), 
        panel.grid.major.y = element_blank(), 
        legend.position = "bottom") +
    facet_grid(~tf(Tcf.status)) +
    ggtitle("Shared vs. unique clonotypes") + ylab("number of clonotypes (fraction per mouse)")
```

The following plots show the same data, just with small tweaks each iteration:

```{r }
clonoFracs <- clono_per_pop[, .N, by = c("Tcf.status","shared_between_pops","Mouse")]
clonoFracs[, total_per_Tcf := sum(N), by = c("Mouse","Tcf.status")]
clonoFracs[, frac_per_Tcf := N/total_per_Tcf]
```

```{r fig.width = 20}
p1 <- ggplot(clonoFracs[shared_between_pops==TRUE],
    aes(x = tf(Tcf.status), y = frac_per_Tcf, color = Mouse)) +
    ggbeeswarm::geom_quasirandom() + 
    ggtitle("Fraction of shared clonotypes", subtitle = "found in at least 2 populations") +
    ylab("shared / (shared + unique)") + xlab("population") +
    theme(panel.grid.minor = element_blank(), 
        panel.grid.major.x = element_blank())

p2 <- ggplot(clonoFracs[shared_between_pops==TRUE],
    aes(x = tf(Tcf.status), y = frac_per_Tcf)) +
    geom_boxplot() + 
    ggbeeswarm::geom_quasirandom(aes(color = Mouse)) +  
    ggtitle("Fraction of shared clonotypes", subtitle = "found in at least 2 populations") +
    ylab("shared / (shared + unique)") + xlab("population") +
    theme(panel.grid.minor = element_blank(), 
        panel.grid.major.x = element_blank())

p3 <- ggplot(clonoFracs[shared_between_pops==TRUE],
    aes(x = tf(Tcf.status), y = frac_per_Tcf)) +
    geom_boxplot() + 
    ggbeeswarm::geom_quasirandom(aes(color = Tcf.status), size = 2, alpha = .8) +  
    ggtitle("Fraction of shared clonotypes", subtitle = "found in at least 2 populations") +
    ylab("shared / (shared + unique)") + xlab("population") +
    scale_color_manual(values = tcfcols) +
    theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank())

p1 + p2 + p3
```


```{r}
clonoFracs[shared_between_pops==TRUE, median(frac_per_Tcf), by = Tcf.status]
```

#### Counting cells {.tabset}

```{r}
tcrCells[, .N, Tcf.status]
```
```{r }
tcfHI_and_panc <- unique(clono_per_allPops[grepl("Tcf1hi.*panc", allPops)]$new_cid)
tcrCells.panc <- tcrCells[Tissue == "panc"]
tcrCells.panc[ , from_pLN.Tcf1hi := ifelse(new_cid %in% tcfHI_and_panc, "pLN.Tcf1hi origin", "not found in pLN Tcf1hi")]
```


There are `r nrow(tcrCells[Tissue == "panc"])` cells in total from the pancreas.
`r nrow(tcrCells[ Tissue == "panc" & new_cid %in% shared_across_pops])` (91%) cells of these have a clonotype that's either found in Tcf1.hi and/*or* Tcf1.lo in the pLN ("shared").
`nrow(tcrCells[ Tissue == "panc" & new_cid %in% shared_across_pops & new_cid %in% tcfHI_and_panc])` (89%) cells have a clonotype that's shared with pLN.Tcf1.hi (and possibly *also* in Tcf1.lo)

In the pLN.Tcf1.hi-population, we have `r nrow(tcrCells[ new_cid %in% tcfHI_and_panc & Tcf.status == "pLN.Tcf1hi"])` (of `r nrow(tcrCells[  Tcf.status == "pLN.Tcf1hi"])` total) cells that have a clonotype that's found in the pancreas, too.

The following Venn diagrams depict the **number of cells** from each population that have clonotypes that are either exclusive to that population or found in any of the other populations, too.

```{r}
clono_in_tcfhi <- unique(clono_per_allPops[grepl("Tcf1hi", allPops)]$new_cid)
clono_in_tcflo <- unique(clono_per_allPops[grepl("Tcf1lo", allPops)]$new_cid)
clono_in_panc <- unique(clono_per_allPops[grepl("panc", allPops)]$new_cid)
```

##### Pancreas cells 

```{r fig.height = 10, fig.width = 10}
gplots::venn(list(
    pancCells_clono_in_Tcf1hi = tcrCells[Tcf.status == "pancreas" & new_cid %in% clono_in_tcfhi]$new_cid,
    pancCells_clono_in_Tcf1lo = tcrCells[Tcf.status == "pancreas" & new_cid %in% clono_in_tcflo]$new_cid,
    pancreas.CELLS = tcrCells[Tcf.status == "pancreas"]$new_cid))
```

##### Tcf1-hi cells

```{r fig.height = 10, fig.width = 10}
gplots::venn(list(
    pLNTcf1hiCells_clono_in_panc = tcrCells[Tcf.status == "pLN.Tcf1hi" & new_cid %in% clono_in_panc]$new_cid,
    pLNTcf1hiCells_clono_in_Tcf1lo = tcrCells[Tcf.status == "pLN.Tcf1hi" & new_cid %in% clono_in_tcflo]$new_cid,
    pLNTcf1hi.CELLS = tcrCells[Tcf.status == "pLN.Tcf1hi"]$new_cid))
```

##### Tcf1-lo cells

```{r fig.height = 10, fig.width = 10}
gplots::venn(list(
    pLNTcf1loCells_clono_in_panc = tcrCells[Tcf.status == "pLN.Tcf1lo" & new_cid %in% clono_in_panc]$new_cid,
    pLNTcf1loCells_clono_in_Tcf1hi = tcrCells[Tcf.status == "pLN.Tcf1lo" & new_cid %in% clono_in_tcfhi]$new_cid,
    pLNTcf1lo.CELLS = tcrCells[Tcf.status == "pLN.Tcf1lo"]$new_cid))

```


#### How many Tcf.hi make it to the pancreas?

Here, I extract all clonotypes that aren't unique; this includes cases where a clonotype might only be found in pLN-Tcf.lo + pancreas:

```{r fig.width = 12}
p1 <- tcrCells[new_cid %in% shared_across_pops, c("Mouse","Tcf.status","pLN_and_panc","new_cid")] %>%
    unique %>%
    ggplot(., aes(x = Mouse, fill = pLN_and_panc)) + 
    geom_bar() + coord_flip() +
    scale_fill_manual(values = c("blue","#D72027")) +
    ylab("number of clonotypes") +
    theme(panel.grid = element_blank())

p2 <- tcrCells[new_cid %in% shared_across_pops, c("Mouse","Tcf.status","pLN_and_panc","new_cid")] %>%
    unique %>%
    ggplot(., aes(x = Mouse, fill = pLN_and_panc)) + 
    geom_bar(position = "fill") + coord_flip() +
    scale_fill_manual(values = c("blue","#D72027")) + 
    ylab("number of clonotypes") +
    theme(panel.grid = element_blank())

pw <- p1 + p2
pw + plot_annotation(title = "Non-unique clonotypes")
```

And here we focus on clonotypes where at least one cell is found in the pLN.Tcf1hi population, more directly addressing the question of how many Tcf1-hi-cells make it to the pancreas.

```{r fig.width = 12}
clono_tcf1hi <- tcrCells[new_cid %in% shared_across_pops & Tcf.status == "pLN.Tcf1hi",
    c("Mouse","Tcf.status","pLN_and_panc","new_cid")] %>% unique 
clono_tcf1hi[, Tcf.hi_and := ifelse(pLN_and_panc == TRUE, "pancreas", "pLN.Tcf1lo_only")]

p1 <- ggplot(clono_tcf1hi, aes(x = Mouse, fill = Tcf.hi_and)) + geom_bar() + coord_flip() +
    scale_fill_manual(values = c("#D72027","#5E3DD8")) +
    ylab("number of clonotypes") + xlab("") +
    theme(panel.grid = element_blank())

p2 <- ggplot(clono_tcf1hi, aes(x = Mouse, fill = Tcf.hi_and)) + geom_bar(position = "fill") + coord_flip() +
    scale_fill_manual(values = c("#D72027","#5E3DD8")) + # ggtitle("Non-unique clonotypes with min. 1 cell in the pLN-Tcf1hi population") +
    ylab("number of clonotypes") + xlab("") +
    theme(panel.grid = element_blank())
    
pw <- p1 + p2
pw + plot_annotation("How often do Tcf1-hi cells that we find in multiple populations end up in the pancreas?", 
    subtitle = "Non-unique clonotypes with min. 1 cell in the pLN-Tcf1hi population")
```

This shows that there is some variability between mice, but in most cases, we see well above 50% of the clonotypes ending up in the pancreas.

We can try to translate that to cell numbers (instead of clonotype numbers) to get a better sense of how many cells likely had their origin in pLN1-Tcf.hi
That also illustrates the point of the expansion of those clonotypes that we actually do find shared. I.e. many of the unique clonotypes in the pancreas are also not expanded.

How many *cells* **in the pancreas** represent clonotypes that we also see in pLN.Tcf1-hi?

```{r}
 table(tcrCells.panc$from_pLN.Tcf1hi)
```

```{r fig.width = 12, fig.height = 8}
p1 <- ggplot(tcrCells.panc, aes(x = Mouse, fill = from_pLN.Tcf1hi)) +
    geom_bar() + coord_flip() +
    theme(panel.grid = element_blank(), legend.title = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = c("peachpuff", "#D72027")) +
    ylab("cells") + xlab("")

p2 <- ggplot(tcrCells.panc, aes(x = Mouse, fill = from_pLN.Tcf1hi)) +
    geom_bar(position = "fill") + coord_flip() +
    theme(panel.grid = element_blank(), legend.title = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = c("peachpuff", "#D72027")) +
    ylab("cells") + xlab("")

pw <-  p1 + p2 
pw + plot_annotation(title = "Where do cells from the pancreas originate from?")
```

And, correspondingly, going back to the clonotypes: Pick the **clonotypes** found in the pancreas and color them based on whether they're also seen in the Tcf1-hi population.

```{r fig.height = 10, fig.width = 13}
clono_panc <- tcrCells.panc[, c("new_cid","Mouse","from_pLN.Tcf1hi")] %>% unique
p1 <- ggplot(clono_panc, aes(x = Mouse, fill = from_pLN.Tcf1hi)) +
    geom_bar() + coord_flip() +
    theme(panel.grid = element_blank(), legend.title = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = c("peachpuff", "#D72027")) +
    ylab("clonotypes") + xlab("")

p2 <- ggplot(clono_panc, aes(x = Mouse, fill = from_pLN.Tcf1hi)) +
    geom_bar(position = "fill") + coord_flip() +
    theme(panel.grid = element_blank(), legend.title = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = c("peachpuff", "#D72027")) +
    ylab("clonotypes") + xlab("")

pw <- p1 + p2 
pw + plot_annotation(title = "Do we find clonotypes from the pancreas in our pLN.Tcf1-hi population?")
```

And focusing just on the clonotypes that we find in both pancreas and pLN, how many of those come from pLN-hi?

```{r fig.height = 10, fig.width = 13}
clono_panc <-  clono_per_pop[grepl("pLN.*panc", allPops), c("new_cid","Mouse","allPops")] %>% unique
clono_panc[, pancreas_and := ifelse(grepl("Tcf1hi", allPops), "pLN.Tcf1hi", "pLN.Tcf1lo")]
p1 <- ggplot(clono_panc, aes(x = Mouse, fill = pancreas_and)) +
    geom_bar() + coord_flip() +
    theme(panel.grid = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = tcfcols) +
    ylab("clonotypes") + xlab("")

p2 <- ggplot(clono_panc, aes(x = Mouse, fill = pancreas_and)) +
    geom_bar(position = "fill") + coord_flip() +
    theme(panel.grid = element_blank(),
        legend.position = "bottom") +
    scale_fill_manual(values = tcfcols) +
    ylab("clonotypes") + xlab("")

pw <- p1 + p2 
pw + plot_annotation(title = "Do we find clonotypes from the pancreas that we also see in the pLN primarily in either Tcf population?", subtitle = "focus on clonotypes found in panc + pLN")
```

## Alluvial plots

The plots below have all clonotypes that are NOT SHARED among different types of cells shown in YELLOW.
All clonotypes that are NOT YELLOW are present in **>1 population of cells**.

```{r alluvials_summary_indCloneColors, fig.height = 25, fig.width = 45}
pl <- lapply(unique(clono_per_pop$Mouse), function(x){
    you <- clono_per_pop[Mouse == x]
    yo.tmp <- clono_per_pop[Mouse ==  x]
    setorder(yo.tmp, -shared_between_pops, -N)
    
    you$Tcf.status <- factor(you$Tcf.status,levels = c("pLN.Tcf1hi","pLN.Tcf1lo","pancreas"), ordered = TRUE)
    
    ## recount 
    you <- you[ , sum(N), by = c("new_cid","Tcf.status") ] %>%
        you[., on = c("new_cid","Tcf.status")]
    you[, total := sum(V1), by = c("Mouse","Tcf.status")]
    you[, frac := V1/total]
    
    you$new_cid <- factor(you$new_cid, levels = unique(yo.tmp$new_cid), ordered = TRUE)
    you$pseudo_clono.ID <- factor(you$pseudo_clono.ID,
        levels = unique(yo.tmp$pseudo_clono.ID), ordered = TRUE)
    
    clon_cols_n <- nlevels(factor(you$pseudo_clono.ID)) -1
    clon_cols <- colorRampPalette(c("#0D0887FF" ,"#7E03A8FF", "#CC4678FF", "#F89441FF"))(clon_cols_n)
    
    P <- ggplot(you, 
        aes(x = Tcf.status, y = frac, stratum = new_cid,
            alluvium = new_cid, fill = pseudo_clono.ID, label = pseudo_clono.ID)) +
        geom_stratum(width = .3, lwd = .1) +
        geom_alluvium( width = .3) + 
        scale_fill_manual(values = c(clon_cols,"khaki1")) +
        theme(legend.position = "none",
            panel.grid.minor=element_blank(),
            panel.grid.major=element_blank()) + 
        ggtitle(x) + ylab("fraction of cells per population") + xlab("")
    return(P)
})
ABCutilities::MultiPlotList(pl, cols = 7)
```

Here are the absolute counts instead of the frequencies:

```{r alluvials_summary_indCloneColors_counts, fig.height = 25, fig.width = 45}
pl <- lapply(unique(clono_per_pop$Mouse), function(x){
    you <- clono_per_pop[Mouse == x]
    yo.tmp <- clono_per_pop[Mouse ==  x]
    setorder(yo.tmp, -shared_between_pops, -N)
    
    you$Tcf.status <- factor(you$Tcf.status,levels = c("pLN.Tcf1hi","pLN.Tcf1lo","pancreas"), ordered = TRUE)
    
    ## recount 
    you <- you[ , sum(N), by = c("new_cid","Tcf.status") ] %>%
        you[., on = c("new_cid","Tcf.status")]

    you$new_cid <- factor(you$new_cid, levels = unique(yo.tmp$new_cid), ordered = TRUE)
    you$pseudo_clono.ID <- factor(you$pseudo_clono.ID,
        levels = unique(yo.tmp$pseudo_clono.ID), ordered = TRUE)
    
    clon_cols_n <- nlevels(factor(you$pseudo_clono.ID)) -1
    clon_cols <- colorRampPalette(c("#0D0887FF" ,"#7E03A8FF", "#CC4678FF", "#F89441FF"))(clon_cols_n)
    
    P <- ggplot(you, 
        aes(x = Tcf.status, y = N, stratum = new_cid,
            alluvium = new_cid, fill = pseudo_clono.ID, label = pseudo_clono.ID)) +
        geom_stratum(width = .3, lwd = .1) +
        geom_alluvium( width = .3) + 
        scale_fill_manual(values = c(clon_cols,"khaki1")) +
        theme(legend.position = "none",
            panel.grid.minor=element_blank(),
            panel.grid.major=element_blank()) +
        ggtitle(x) + ylab("number of cells per population") + xlab("")
    return(P)
})
ABCutilities::MultiPlotList(pl, cols = 7)
```

# SessionInfo

```{r}
sessionInfo()
```


